---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "90%" # ,
  # dev = "svg"
)
```

# QATS (Quick Adaptive Ternary Segmentation)

<!-- badges: start -->

<!-- badges: end -->

In hidden Markov models (HMM), one observes a noisy version of an unobservable finite-state Markov chain. One of the main components of HMM is to decode that signal, that is to estimate the sequence of hidden states at the origin of the observation sequence. Existing decoding algorithms such as [Viterbi (1967)](https://doi.org/10.1109/TIT.1967.1054010) algorithm have computational complexity at best linear in the size of the observed sequence, and sub-quadratic in the size of the state space.

Assuming that the observation sequence is stored as specific cumulative sums, we present Quick Adaptive Ternary Segmentation (QATS), a procedure which decodes the hidden sequence in polylogarithmic computational complexity in the size of the sequence, and cubic in the size of the state space. In essence, the estimated sequence of states sequentially maximizes local likelihood scores among all local paths with at most three segments. The latter search is performed only approximately using an adaptive search procedure. The resulting sequence is admissible in the sense that all transitions occur with positive probability.

QATS and Viterbi algorithm, as well as algorithms to compute the generalized risk-based estimator of [Lember and Koloydenko (2014)](https://dl.acm.org/doi/10.5555/2627435.2627436) and the K-segment approach of [Titsias, Holmes and Yau (2016)](https://doi.org/10.1080/01621459.2014.998762) are implemented in both **R** and **C++**.

## Installation

You can install the development version of **QATS** from [GitHub](https://github.com/) with:

```{r, eval=FALSE}
install.packages("devtools")
devtools::install_github("AlexandreMoesching/QATS")
```

## Usage

This is a basic example which shows how to solve a common problem:

```{r}
# Load package
library(QATS)

# Set any seed, just for the example...
set.seed(123)

# Cardinality of the space of hidden states
m <- 5

# Standard deviation of the normal emission distributions
sigma <- 0.5

# Size of the observation sequence
n <- 1e3 + 1

# Expected number of change points
K <- 7

# Generate a sequence
par <- sample.HMM(
  n = n, m = m, K = K,
  emi.dist = "normal",
  emi.param = list(sigma = sigma)
)

# Extract the hidden (xx.0) and observed (yy) sequences
xx.0 <- par$xx
yy <- par$yy

# Actual number of change points of the generated sequence:
sum(diff(xx.0) != 0)
```

The function `sample.HMM()` already pre-computes useful quantities for later such as the log-probability transition matrix, log-initial distribution, densities and log-densities evaluated at the observations, cumulative sums of log-densities, etc. The list of parameters will then be passed to the different functions.

We can now plot the generated hidden sequence (black line) and observations (grey points):

```{r}
display.result(xx.0, par = par, yy = yy)
```

To use Viterbi algorithm in order to estimate the hidden sequence, simply do:

```{r}
res <- Viterbi.CPP(par)
xx.1 <- res$xx
(tt <- as.vector(res$time)) # Time in seconds
```

To use the PMAP algorithm in order to estimate the hidden sequence, simply do:

```{r}
res <- PMAP.CPP(par)
xx.1.PMAP <- res$xx
(tt <- as.vector(res$time)) # Time in seconds
```

To use QATS, run:

```{r}
res <- QATS.CPP(par)
xx.2 <- res$xx
(tt <- as.vector(res$time)) # Time in seconds
```

Here is a plot of the true path (black), Viterbi-path (blue) and QATS-path (red):

```{r}
display.result(xx.0, xx.1, xx.2, par)
```

The following line shows that Viterbi and QATS produce the same path, and that both did not estimate correctly 5 states out of the 1001.

```{r}
c(sum(xx.0 != xx.1), sum(xx.1 != xx.1.PMAP), sum(xx.1 != xx.2))
```

## Comparison of QATS, Viterbi, risk-based estimators and K-segmentation

```{r}
# Set a larger observation sequence, the other parameters remain unchanged
n <- 1e6 + 1

# Generate a sequence
par <- sample.HMM(
  n = n, m = m, K = K,
  emi.dist = "normal",
  emi.param = list(sigma = sigma)
)
xx.0 <- par$xx
sum(diff(xx.0) != 0)

# 1. Viterbi
res <- Viterbi.CPP(par)
xx <- matrix(res$xx, nrow = 1)
tt <- as.vector(res$time)
rownames(xx)[1] <- names(tt)[1] <- "Viterbi"

# 2. Pointwise MAP - manual
res <- PMAP.CPP(par)
xx <- rbind(xx, res$xx)
tt <- c(tt, as.vector(res$time))
rownames(xx)[length(tt)] <- names(tt)[length(tt)] <- "pMAP-man"

# 3. Pointwise MAP
res <- G_classifier.CPP(par, 1, 0, 0, 0)
xx <- rbind(xx, res$xx)
tt <- c(tt, as.vector(res$time))
rownames(xx)[length(tt)] <- names(tt)[length(tt)] <- "pMAP"

# 4. Maximum prior probability
res <- G_classifier.CPP(par, 0, 0, 0, 1)
xx <- rbind(xx, res$xx)
tt <- c(tt, as.vector(res$time))
rownames(xx)[length(tt)] <- names(tt)[length(tt)] <- "MPP"

# 5. Marginal prior mode
res <- G_classifier.CPP(par, 0, 0, 1, 0)
xx <- rbind(xx, res$xx)
tt <- c(tt, as.vector(res$time))
rownames(xx)[length(tt)] <- names(tt)[length(tt)] <- "MPM"

# 6. Generalized Viterbi
res <- G_classifier.CPP(par, 0, 1, 0, 1)
xx <- rbind(xx, res$xx)
tt <- c(tt, as.vector(res$time))
rownames(xx)[length(tt)] <- names(tt)[length(tt)] <- "gViterbi"

# 7. K-segmentation
K_max <- K + 4
res <- K_segmentation.CPP(par, K_max)
xx <- rbind(xx, res$xx)
tt <- c(tt, rep(as.vector(res$time), K_max))
rownames(xx)[length(tt) - ((K_max - 1):0)] <-
  names(tt)[length(tt) - ((K_max - 1):0)] <-
  paste(rep("K-seg", K_max), 1:K_max)

# 8. QATS
res <- QATS.CPP(par)
xx <- rbind(xx, res$xx)
tt <- c(tt, as.vector(res$time))
rownames(xx)[length(tt)] <- names(tt)[length(tt)] <- "QATS"

# Compute errors
n.estim <- nrow(xx)
fit_eval <- matrix(0, nrow = n.estim, ncol = 4)
colnames(fit_eval) <- c("l0", "l1", "l2", "V-Measure")

for (w in 1:n.estim) {
  fit_eval[w, 1] <- lp_norm(xx.0, xx[w, ], 0)
  fit_eval[w, 2] <- lp_norm(xx.0, xx[w, ], 1)
  fit_eval[w, 3] <- lp_norm(xx.0, xx[w, ], 2)
  fit_eval[w, 4] <- V_measure(xx.0, xx[w, ])
}
rownames(fit_eval) <- names(tt)

cbind(fit_eval, tt)
```

## Step-by-step computation of QATS

For that, make sure that the sample size is not too large ($n < 10^3$).

```{r, eval=FALSE}
set.seed(1234)
n <- 3e2 + 1
par <- sample.HMM(
  n = n, m = m, K = K,
  emi.dist = "normal",
  emi.param = list(sigma = sigma)
)
res.Vit <- Viterbi.CPP(par)
res.QATS <- QATS.display(par$xx, res.Vit$xx, par)
```
